Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-18 paper-1/manuscript.tex
Separate T and theta

When a neuron's membrane potential reached the firing threshold $V_\theta$, a spike was emitted, and the potential was clamped to $V_m = V_{reset} = -74$~mV for a fixed refractory period of 3~ms (excitatory neurons). Inhibitory neurons were modeled as fast-spiking cells with a refractory period of 2~ms and a constant firing threshold $V_\theta = \theta_0 = -54$~mV \citep{Mensi2012-au}. In contrast, excitatory neurons were modeled with threshold adaptation (TA) as in \cite{Teeter2018-iz}, following
\begin{align}
    V_\theta &= \theta_0 + \T{}{}(t) \nonumber \\
    \tau_{\theta} \frac{d\T{}{}}{dt} &= -\T{}{} + \hat{\theta} \delta(t - \hat{t}) \label{eqn-TA}
\end{align}
with increment $\hat{\theta} = 1$~mV and decay time constant $\tau_{\theta} = 1$~s \citep{Pozzorini2015-ei}.

@@

Throughout this paper, we refer to three main quantities of interest: the stimulus response $\R{}{}$ defined in the next paragraph, the threshold adaptation value $\T{}{}$ defined in Equation \ref{eqn-TA}, and the short-term depression effect $\D{}{}$, which will be introduced later in Equation \ref{eqn-D}. We employ sub- and superscripts to identify the precise quantity as follows, using $\R{}{}$ for illustration: $\R[seq]{X}{p}$ refers to the response of neuron(s) $p$, averaged across all trials of type $X$ in the sequence $seq$. Where $p$ does not identify individual neurons (e.g., $i$ or $j$), but rather a subpopulation (e.g., $exc$) or the whole network (no index), this refers to the median (of the individual quantity $\R[seq]{X}{i}$) across the indicated set of neurons. The trial type $X$ is either $A$ or $B$, referring to trials with stimulation at these sites, or $N$, indicating all \textbf{n}on-target trials B through E. Finally, we use the notation $\Delta \R{}{}$ to refer to the sequence contrast $\R[dev]{}{} - \R[msc]{}{}$, and $\mean{\R{}{p}}$ to indicate the mean, rather than median, of the individual quantities in a population $p$.

We define the response $\R{X}{i}$ as the mean spike count of neuron $i$ across time bins in trials of type $X$. Almost equivalently, we define $\T{X}{i}$ (and $\D{X}{i}$ analogously) as the mean across trials of type $X$ of $\T{}{i}(t=0)$, i.e., the average TA voltage at the beginning of trials, immediately before stimulation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower $\T{A}{}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence ($\R[dev]{A}{i}$) and \msc{} sequence ($\R[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \R{A}{i}$).
\textbf{b} Target trial average of the threshold adaptation voltage $\T{}{}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\T{}{}$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\T{}{}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\T{}{}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\T{}{}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\T{}{}$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\T{}{}$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\T{}{}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast in $\T{}{}$ at the start of target trials, i.e., $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$ corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\T{}{}$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\T{}{}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $\T{}{}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $\T{}{}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{}{}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $\T{}{}$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

@@

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in all sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\T{}{}$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, Figure \ref{fig6}c) and the median contrast in $\T{}{}$ at the start of B trials ($\Delta \T{B}{}$, Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

>>>>

When a neuron's membrane potential reached the firing threshold $V_\theta$, a spike was emitted, and the potential was clamped to $V_m = V_{reset} = -74$~mV for a fixed refractory period of 3~ms (excitatory neurons). Inhibitory neurons were modeled as fast-spiking cells with a refractory period of 2~ms and a constant firing threshold $V_\theta = \theta_0 = -54$~mV \citep{Mensi2012-au}. In contrast, excitatory neurons were modeled with threshold adaptation (TA) as in \cite{Teeter2018-iz}, following
\begin{align}
    V_\theta &= \theta_0 + \theta(t) \nonumber \\
    \tau_{\theta} \frac{d\theta}{dt} &= -\theta + \hat{\theta} \delta(t - \hat{t}) \label{eqn-TA}
\end{align}
with increment $\hat{\theta} = 1$~mV and decay time constant $\tau_{\theta} = 1$~s \citep{Pozzorini2015-ei}.

@@

Throughout this paper, we refer to three main quantities of interest: the stimulus response magnitude $\R{}{}$, the threshold adaptation value $\T{}{}$, and the short-term depression value $\D{}{}$. We define the response $\R{X}{i}$ as the mean number of spikes fired by neuron $i$ across all trials of type $X$, i.e., the average response to $X$. We define $\T{X}{i}$ as the mean across trials of type $X$ of $\theta_i(t=0)$, i.e., the average TA voltage at the beginning of trials, immediately before stimulation. Lastly, the STD quantity $\D{X}{i}$ will be introduced later in Equation \ref{eqn-D}.

We employ sub- and superscripts to identify the precise quantity as follows, using $\R{}{}$ for illustration: $\R[seq]{X}{p}$ refers to the response of neuron(s) $p$, averaged across all trials of type $X$ in the sequence $seq$. We use $p$ to either identify individual neurons (e.g., $i$ or $j$), or a subpopulation (e.g., $exc$) or the whole network (no index), in which case the designated quantity is the median (of the individual quantities $\R[seq]{X}{i}$) across the indicated set of neurons. The trial type $X$ is either $A$ or $B$, referring to trials with stimulation at these sites, or $N$, indicating all \textit{n}on-target trials B through E. Finally, we use the notation $\Delta \R{}{}$ to refer to the sequence contrast $\R[dev]{}{} - \R[msc]{}{}$, and $\mean{\R{}{p}}$ to indicate the mean, rather than median, of the individual quantities in a population $p$.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower initial $\theta$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence ($\R[dev]{A}{i}$) and \msc{} sequence ($\R[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \R{A}{i}$).
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
\textbf{e} $\Delta \T{A}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \T{A}{i}$ vs $\Delta \R{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\theta$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\theta$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We quantified the contrast in $\theta$ at the start of target trials as $\Delta \T{A}{i} = \T[dev]{A}{i} - \T[msc]{A}{i}$, where $\T{}{i} = \theta_i(t=0)$, corresponding to the leftmost datapoints of the contrast histogram in Figure \ref{fig4}b, and correlated this with the target trial response size contrast $\Delta \R{A}{i}$, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median of $\T{A}{i}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the correlation across excitatory neurons between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\T[dev]{A}{i}$ tended to show a greater increase in their response to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\T[dev]{}{}$ is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\T{A}{i}$, i.e., $\theta_i$ at the start of A trials (mean over trials), in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
\textbf{d} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $\theta$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

@@

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in all sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\theta$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, Figure \ref{fig6}c) and the median contrast in $\T{B}{}$ ($\Delta \T{B}{}$, Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).
